Simple repression draft

### 2020-10-22 doc/resubmission/appendix_03_bayesian_inference.tex
final pass before resubmission

\subsubsection{Prior selection}
The different models explored in this work embraced different levels of
coarse-graining that resulted in a diverse number of parameters for different
models. For each of these model configurations Bayes' theorem demands from us to
represent our preconceptions on the possible parameter values in the form of the
prior $P(\vec{\theta})$. Throughout this work for models with $> 1$ parameter we
assign independent priors to each of the parameters; this is
\begin{equation}
P(\vec{\theta}) = \prod_{i=1}^n P(\theta_i).
\end{equation}
Although it is not uncommon practice to use non-informative, or maximally
uninformative priors, we are of the mindset that this is a disservice to the
philosophical and practical implications of Bayes' theorem. It sounds almost
contradictory to claim that can we represent our thinking about a natural
phenomenon in the form of a mathematical model --in the context of Bayesian
inference this means choosing a form for the likehihoods, and even making
this choice presupposes prior understanding or assumptions as to the
relevant features in the system under study-- but that we have absolutely
no idea what the parameter values could or could not be. We therefore make
use of our own expertise, many times in the form of order-of-magnitude
estimates, to write down weakly-informative prior distributions
for our parameters.

@@

Throughout this Appendix we will appeal to the convenient notation for
probability distributions of the form
\begin{equation}
m \sim \text{Poisson}(\lambda),
\end{equation}
where the simbol ``$\sim$'' can be read as \textit{is distributed according to}.
So the previous equation can be read as: the mRNA copy number $m$ is distributed
according to a Poisson distribution with parameter $\lambda$. Our objective then
is to compute the posterior probability distribution $P(\lambda \mid D)$, where,
as in the main text, $D = \{ m_1, m_2, \ldots, m_N \}$ are the data consisting
of single-cell mRNA counts. Since we can assume that each of the cells mRNA
counts are independent of any other cells, our likelihood function $P(D \mid
\lambda)$ consists of the product of $N$ Poisson distributions.

@@

The structure of the data in~\cite{Jones2014} for regulated promoters tuned
these two parameters independently. In their work the production of the LacI
repressor was under the control of an inducible promoter regulated by the TetR
repressor as schematized in Figre~\ref{figS:aTc_circuit}. When TetR binds to the
small molecule anhydrotetracycline (aTc), it shifts to an inactive conformation
unable to bind to the DNA. This translates into an increase in gene expression
level. In other words, the higher the concentration of aTc added to the media,
the less TetR repressors that can control the expression of the \textit{lacI}
gene, so the higher the concentration of LacI repressors in the cell. So by 
tuning the amount of aTc in the media where the experimental strains were grown
they effectively tune $k^+$ in our simple theoretical model. On the other hand
to tune $k^-$ the authors swap three different binding sites for the LacI 
repressor, each with different repressor-DNA binding affinities previously 
characterized \cite{Garcia2011a}.

@@

Formally now, denote the set of seven repressor rates to be inferred as
\begin{equation}
\vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\}.
\end{equation}
Note that since the repressor copy numbers are not known directly as explained
before, we label their association rates by the concentration of aTc. Bayes
theorem reads simply
\begin{equation}
p(\vect{k}, k_i, b \mid D)
\propto
p(D \mid\vect{k}, k_i, b) p(\vect{k}, k_i, b),
\end{equation}
where $D$ is the set of all $N$ observed single-cell mRNA counts across the
various conditions. We assume that individual single-cell measurements are
independent so that the likelihood factorizes as
\begin{equation}
p(D \mid\vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid \vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid k_j^+, k_j^-, k_i, b)
\end{equation}
where $k_j^\pm$ represent the appropriate binding and unbinding rates for the
$j$-th measured cell. Our likelihood function, previously derived in
Appendix~\ref{sec:gen_fcn_appdx}, is given by the rather complicated result in
Eq.~\ref{eq:p_m_bursty+rep_appdx}, which for completeness we reproduce here as
\begin{equation}
\begin{split}
p(m \mid k_R^+, k_R^-, k_i, b) = & ~\frac{
        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
        }
        {
        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
        }
\frac{b^m}{m!}
\\
&\times {_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b).
\end{split}
\label{eq:p_m_bursty+rep_infreprint}
\end{equation}
where $\alpha$ and $\beta$, defined for notational convenience, are
\begin{align}
\begin{split}
\alpha &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\beta &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
\end{split}
\end{align}

@@

\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{../../figures/si/figS0X_histograms.pdf}
\caption{\textbf{Theory-experiment comparision of mRNA distributions for
regulated promoters.} Comparison of the inference (color shaded area
encloses the 95\% of all posterior predictive check samples while the color
solid line represents the median of all samples) vs the experimental
measurements (black lines) for different regulated promoters with different
operators (columns) and aTc concentrations (rows) from
Ref.~\cite{Jones2014}.}
\label{figS:reg_histograms}
\end{figure}

>>>>

\subsubsection{Prior selection}
The different models explored in this work embraced different levels of
coarse-graining that resulted in a diverse number of parameters for different
models. For each of these model configurations Bayes' theorem demands from us to
represent our preconceptions on the possible parameter values in the form of the
prior $P(\vec{\theta})$. Throughout this work for models with $> 1$ parameter we
assign independent priors to each of the parameters; this is
\begin{equation}
P(\vec{\theta}) = \prod_{i=1}^n P(\theta_i).
\end{equation}
Although it is not uncommon practice to use non-informative, or maximally
uninformative priors, we are of the mindset that this is a disservice to the
philosophical and practical implications of Bayes' theorem. It sounds almost
contradictory to claim that can we represent our thinking about a natural
phenomenon in the form of a mathematical model --in the context of Bayesian
inference this means choosing a form for the likelihoods, and even making
this choice presupposes prior understanding or assumptions as to the
relevant features in the system under study-- but that we have absolutely
no idea what the parameter values could or could not be. We therefore make
use of our own expertise, many times in the form of order-of-magnitude
estimates, to write down weakly-informative prior distributions
for our parameters.

@@

Throughout this Appendix we will appeal to the convenient notation for
probability distributions of the form
\begin{equation}
m \sim \text{Poisson}(\lambda),
\end{equation}
where the symbol ``$\sim$'' can be read as \textit{is distributed according to}.
So the previous equation can be read as: the mRNA copy number $m$ is distributed
according to a Poisson distribution with parameter $\lambda$. Our objective then
is to compute the posterior probability distribution $P(\lambda \mid D)$, where,
as in the main text, $D = \{ m_1, m_2, \ldots, m_N \}$ are the data consisting
of single-cell mRNA counts. Since we can assume that each of the cells mRNA
counts are independent of any other cells, our likelihood function $P(D \mid
\lambda)$ consists of the product of $N$ Poisson distributions.

@@

The structure of the data in~\cite{Jones2014} for regulated promoters tuned
these two parameters independently. In their work the production of the LacI
repressor was under the control of an inducible promoter regulated by the TetR
repressor as schematized in Figure~\ref{figS:aTc_circuit}. When TetR binds to the
small molecule anhydrotetracycline (aTc), it shifts to an inactive conformation
unable to bind to the DNA. This translates into an increase in gene expression
level. In other words, the higher the concentration of aTc added to the media,
the less TetR repressors that can control the expression of the \textit{lacI}
gene, so the higher the concentration of LacI repressors in the cell. So by 
tuning the amount of aTc in the media where the experimental strains were grown
they effectively tune $k^+$ in our simple theoretical model. On the other hand
to tune $k^-$ the authors swap three different binding sites for the LacI 
repressor, each with different repressor-DNA binding affinities previously 
characterized \cite{Garcia2011a}.

@@

Formally now, denote the set of seven repressor rates to be inferred as
\begin{equation}
\vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\}.
\end{equation}
Note that since the repressor copy numbers are not known directly as explained
before, we label their association rates by the concentration of aTc. Bayes
theorem reads simply
\begin{equation}
p(\vect{k}, k_i, b \mid D)
\propto
p(D \mid\vect{k}, k_i, b) p(\vect{k}, k_i, b),
\end{equation}
where $D$ is the set of all $N$ observed single-cell mRNA counts across the
various conditions. We assume that individual single-cell measurements are
independent so that the likelihood factorizes as
\begin{equation}
p(D \mid\vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid \vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid k_j^+, k_j^-, k_i, b)
\end{equation}
where $k_j^\pm$ represent the appropriate binding and unbinding rates for the
$j$-th measured cell. Our likelihood function, previously derived in
Appendix~\ref{sec:gen_fcn_appdx}, is given by the rather complicated result in
Eq.~\ref{eq:p_m_bursty+rep_appdx}, which for completeness we reproduce here as
\begin{equation}
\begin{split}
p(m \mid k_R^+, k_R^-, k_i, b) = & ~\frac{
        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
        }
        {
        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
        }
\frac{b^m}{m!}
\\
&\times {_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b).
\end{split}
\label{eq:p_m_bursty+rep_infreprint}
\end{equation}
where $\alpha$ and $\beta$, defined for notational convenience, are
\begin{equation}
\begin{split}
\alpha &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\beta &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
\end{split}
\end{equation}

@@

\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{../../figures/si/figS0X_histograms.pdf}
\caption{\textbf{Theory-experiment comparison of mRNA distributions for
regulated promoters.} Comparison of the inference (color shaded area
encloses the 95\% of all posterior predictive check samples while the color
solid line represents the median of all samples) vs the experimental
measurements (black lines) for different regulated promoters with different
operators (columns) and aTc concentrations (rows) from
Ref.~\cite{Jones2014}.}
\label{figS:reg_histograms}
\end{figure}
